---
title: "Impact of severe weather conditions in the USA"
author: Pawel Daniluk
date: 2016-11-03
output:
  html_document
---

## Synopsis
This report summarizes analysis of impact of severe weather conditions on public
health and property damage caused by severe weather conditions.

The analysis is based on U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, as of November 2011.

The dataset is available for [download](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). 

Complete [documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) for the research dataset is published and available for download.

As described in documentation, the dataset contains detailed records of:

- Storms and other significant weather phenomena, having sufficient intensity to cause loss of life, injuries, significant property damage, and/or disruption to commerce;

- Rare, unusual, weather phenomena that generate media attention, such as snow flurries in South Florida or the San Diego coastal area; and

- Other significant meteorological events, such as record maximum or minimum temperatures or precipitation that occur in connection with another event. 

## Loading raw data

Compressed data set is downloaded and read into data frame object. For improved performance, data data is being prevented from downloading if already present in the working directory. Decompressed data is stored in working directory as RDat file, to be loaded into memory as needed. 

```{r, get_data}
if (!dir.exists("./data/")) {

    dir.create("./data/")
    # Downloading data and documentation
    surl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    durl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"
    furl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf"

    download.file(surl, "./data/storm.bz2", mode="wb")
    download.file(durl, "./data/data_doc.pdf", mode="wb")
    download.file(furl, "./data/data_faq.pdf", mode="wb")

    # Read compressed file into data frame
    # read.csv function will decompress bz2 archive

    storm <- read.csv("./data/storm.bz2")

    # Save decompressed data set to RDat. file for reuse.
    save(storm, file="./data/storm.RDat")
}

# Load storm data from RDat. file if not present in memory
if(!exists("storm")) 
    load("./data/storm.RDat")
```
## Data description
```{r, data_props, echo=FALSE}
dims <- dim(storm)
state_lev_no <- length(levels(storm$STATE))
evtype_lev_no <-  length(levels(storm$EVTYPE))
``` 
The analysis set contains severe weather events, recorded between the year 1950 and November 2011. The Storm database contains `r dims[1]` rows and `r dims[2]` columns. The colums used for analysis are listed in the table below.

| Column name 	| Type    	                    | Description      	|
|-------------	|------------------                 |------------------	|
| EVTYPE      	| Factor: `r evtype_lev_no` levels  | Type of weather event          |
| FATALITIES  	| Numeric      	                    | Number of fatalities per event |
| INJURIES     	| Numeric    	                    | Number of casualties per event |
| PROPDMG       | Numeric                           | Value of property damage per event |
| CROPDMG      	| Numeric                           | Value of crop damage per event |

Table: Description of selected variables

Since there are `r state_lev_no` levels of EVTYPE varialbe, the output  of the analysis must be limited to be readable. Therefore only top 20 most severe categories will be reported.

## Research methodology
This section describes assumptions made in the analysis.

### Severity of health impact
In order to measure impact of severe weather conditions on both population health and property damage, some measures must be established. Since both fatalities and injuries can be considered harmfull in regard to public health, those two variables should be included. 

However, in the assessment of severity of extrene weather confition in regard to pulic health, fatalities and injuries should not be included with the same weights. For purpouse of this analysis, arbitraty weight of 20% is assumed, i.e. one fatality corresponds to five fatalities in assesing the event severity. Finally, a "CASUALTIES" measure is calculated as follows:
```
CASUALTIES = FATALITIES+.2*INJURIES
```
In addition, it might be informative, to calculate number of casualties per event to measure how destructive is each type of events.

### Severity of economic impact
In order to calculate severity of extreme weather event in terms of economic impact impact, both property damage and crop losses should be taken into account. Since crop damage and property damage are two different categories, they will be analyzed separately, without attempting to aggregate the totals, as in case of health impact.

The rationale behind this approach is that crop damage can be extensive in overall damage, but relatively small in monetary terms. 

## Data transformations
In order to obtain severity measures calculated as described in the previous sections, the **dplyr** package is used.
Firstly, the measures of health impact are calculated. The results are arranged in regard to casualties count.
```{r, load_data, echo=FALSE, warning=FALSE, message=FALSE}
# Load data to dplyr table for transformation
# Loading required packages (dplyr, tidyr, ggplot2)
library(dplyr)
library(tidyr)
library(ggplot2)
storm <- tbl_df(storm)
```

```{r, cas_data_trans}
# Calculation of casualty value as fatalities + 0.2*injuries (arbitrary)
storm_cas <- storm %>%
    mutate(casualties = FATALITIES + 0.2*INJURIES
           , casualties=ceiling(casualties)
           )

# Calculating casualty values by event type
cas_evtype <- storm_cas %>%
    group_by(EVTYPE) %>%
    summarise(cas_sum = sum(casualties)
              , fat_sum = sum(FATALITIES)
              , inj_sum = sum(INJURIES)
              , ev_n = n()
              , cas_per_event = cas_sum/n()) %>%
    arrange(desc(cas_sum)) 
```
In order to draw required barplot, data frame need to be transposed with **tidyr** package function **gather**. 
Top 5 
```{r, cas_gather}
cas_bar <- head(cas_evtype,5) %>% 
  rename(Casualties=cas_sum, Fatalities=fat_sum, Injuries=inj_sum) %>%
  select(-c(ev_n, cas_per_event)) %>%
  gather(Label, value, -EVTYPE)
``` 

The property and crop damage will be calculated in the similar fashion, as shown in the following code listing.

```{r, dmg_data_trans}

```

## Assessment of impact on public health 
The top most severe types of weather events in respect to public health are listed in the following table. :

Printed with knirt:kable function to transform r output into markdown table.
```{r health_table}
# supress scientific notation in printed output
options(scipen=999)

library(knitr)
cas <- head(cas_evtype, 20) 
names(cas) <- c("Event type", "Total casualties", "Total Fatalities", "Total injuries"
    , "Number of events", "Casualty rate")
kable(cas,  digits=2, caption = "Top 20 event types most severe to public health")
```

Based on the gathered data, it appears, that Tornadoes are leading cause of casualties in the USA, resulting in overall `r cas[1,2]` fatalities, and `r cas[1,3]` injuries. The calculated casualties amounted to `r cas[1,4]`. It is interesting to note however, that although Tornadoes results in highest number of casualties, the actual casualty rate is modest `r cas[1,6]`. High position of Tornadoes result from high number of this type of events.  

Interestingly, the most destructive event types, such as HURRICANE/TYPHOON (3.68 casualty rate), Heat and Excessive Heat (1.95 and 1.79 respectively) has relatively low impact on public health, because of their low frequency of occurrence.

Finally, the results can be graphically presented in form of bar plot, created with **ggplot2** package.
```{r, cas_plot, warning=FALSE, message=FALSE}
# Transform data to horizontal format to produce barplot
cas2 <- head(cas_evtype, 5)
library(tidyr)

  
# Barplot for top 5 event types

library(ggplot2)
ggplot(cas_bar, aes(x=EVTYPE, y=value,fill=Label)) +
  labs(x="Event type", y="Sum", title="Severity of event type (Public Health)") + 
  geom_bar(stat="identity", position="dodge")
```


## Assessment of property impact
